---
title: "Alternating Minimization Allgorithms (ALS) and other alternatives"
author: "Khaled Fouda"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = F, cache = FALSE)
library(kableExtra)
library(magrittr)
library(tidyverse)
knitr::opts_knit$set(root.dir="/mnt/campus/math/research/kfouda/main/HEC/Youssef/HEC_MAO_COOP/")
```


```{r include=TRUE, message=FALSE, warning=FALSE}
source("./code_files/Mao_import_lib.R")
data_dir = "./Mao/saved_data/"

```

## References:

[1] Hastie (2015) Matrix Completion and Low-Rank SVD via Fast Alternating Least Squares
[2]



Muilding on our model in the document "Soft Impute With Covariates", we explore alternative ways of computing the low-rank matrix "M" while achieving the same results. The alternative ways are much faster that they outperform Mao's model.

We begin by introducing the ALS algorithm for estimating M, first without covariates:

Consider our usual model (model 8 in Soft Impute With Covariates):

$$
\begin{aligned}
minimize_M && \frac{1}{2} \|Y-M\|^2_F + \lambda ||M||_*
\end{aligned}
$$
which as the following soluion



$$
\begin{aligned}
\hat{M} = S_\lambda(Y) = U D_\lambda V'&& with && D_\lambda = diag[(d_1-\lambda)_+\cdots(d_r-\lambda)_+]
\end{aligned}
$$
where $r$ acts as an upper bound to the rank of $\hat{M}$. Since we don't require the full SVD decomposition to compute $\hat{M}$ and we only need the rows/columns upto $r$, the authors have discussed the possibility of using alternative methods to do partial svd decomposition. Specifically, they discussed the use of the algorithm "PROPABK" which is an state-of-the-art method of computing SVD decomposition upto certain number of Eigen Values. However, they didn't do it in their implementation and they used full SVD decomposition instead. 


A second trick they proposed to avoid the high computations is that they rewrote $\hat{Y}$ as 

$$
\begin{aligned}
\hat{Y} = [P_\Omega(Y) - P_\Omega(\hat{M})]  + \hat{M}
\end{aligned}
$$

$\hat{Y}$ ubove refers to the filled matrix where observed values are left as they are and missing values are filled from $\hat{M}$. The first term above is a very sparse matrix since the missing values are left as 0. The second term is not sparse but is low-rank. Using the two terms instead of $\hat{Y}$ should improve the speed of matrix operations. Again, they proposed it but didn't use it. I attempt to use it but it provided no significant improvement over using $\hat{Y}$ as is. Maybe this will be different for extremely large matrices as I tested with dimensions up-to $1000\times 1000$.

Hastie et al., and based on the approach by Rennie and Srebero (2005), they considered the following model instead


$$
\begin{aligned}
minimize_{A,B} && \frac{1}{2} \|Y-AB^T\|^2_F + \frac{\lambda}{2} (||A||^2_F+||B||^2_F)  && (6)
\end{aligned}
$$
Where $A$ is $n_1\times r$ and $B$ is $n_2 \times r$ and $M=AB^T$. The loss above is called maximum-margin matrix factorization (MMMF) which is not convex in A and B, but is bi-convex -- for fixed B, the loss is convex in A and vice-versa. If the solution to our problem has a rank $q \le r$ then the solution to this problem is 


$$
\begin{aligned}
\hat{A} &= U_r S_\lambda(D_r)^{\frac{1}{2}} && (7)\\
\hat{B} &= V_r S_\lambda(D_r)^{\frac{1}{2}}, &&
\end{aligned}
$$

This method could still make use of the sparse + low-rank property. 

Lemma 1 of theorem 2 states the following:

$$
\| M\|_* = min_{A,B:M=AB'} \frac{1}{2}(\|A\|^2_F+\|B\|^2_F)
$$

Suppose we have current estimates of $A$ and $B$ and we wish to compute the new $\tilde{B}$. We will use a filled $Y^*$ using the estimates from the previous iterations and

$$
\begin{aligned}
minimize_{B} && \frac{1}{2} \|Y^*-A\tilde B^T\|^2_F + \lambda ||\tilde B||^2_F.
\end{aligned}
$$
With $Y^* =  P_{\Omega}(Y) + P_{\overline \Omega}(AB') = (P_{\Omega}(Y) + P_{\Omega}(AB')) + AB' $

Which is again, a sparse + low-rank.




*************************************************

**Algorithm 3.1 Rank-Restricted Efficient Maximum-Margin Matrix Factorization: softImpute-ALS**

1. Initialize $A = UD$ where $U_{n_1 \times r}$ is a randomly chosen matrix with orthonormal columns
   and $D = I_r$, the $r \times r$ identity matrix, and $B = VD$ with $V = 0$. Alternatively, any
   prior solution $A = UD$ and $B = VD$ could be used as a warm start.
2. Given $A = UD$ and $B = VD$, approximately solve the following to update $B$:

   $$minimize_B \frac{1}{2} \|P_{\Omega}(Y) - AB^T\|_F^2 + \frac{\lambda}{2} \|B\|_F^2$$

   We achieve that with the following steps:
   - (a) Let $Y^* = (P_{\Omega}(Y) - P_{\Omega}(AB^T)) + AB^T$, stored as sparse plus low-rank.
   - (b) Solve the optimization problem:

     $$minimize_M \frac{1}{2} \|Y^* - AB^T\|_F^2 + \frac{\lambda}{2} \|B\|_F^2$$

     with the solution:

     $$B^T = (D^2 + \lambda I)^{-1}DU^TY^*$$
     $$= (D^2 + \lambda I)^{-1}DU^T(P_{\Omega}(Y) - P_{\Omega}(AB^T))$$
     $$+ (D^2 + \lambda I)^{-1}D^2B^T.$$

   - (c) Use this solution for $\bar{B}$ and update $V$ and $D$:
     1. Bompute the SVD decomposition $\bar{B}D = \tilde{U}\tilde{D}\tilde{V}^T$.
     2. Update $V \leftarrow \tilde{U}$, and $D \leftarrow \tilde{D}$.

3. Given $B = VD$, solve for $A$. My symmetry, this is equivalent to step 2, with $Y^T$ replacing $Y$, and $B$ and $A$ interchanged.
4. Repeat steps (2)-(3) until convergence.
5. Bompute $M = Y^*V$, and then its SVD: $M = U_{\sigma}D_{\sigma}R^T$. Then output $U, V \leftarrow VR$ and $D_{\lambda} = \text{diag}(\sigma_1 - \lambda, \ldots, \sigma_r - \lambda)_+$.

*****************************************************************

**In other words, at iteration i,** we have previous estimates of $A=UD$ and $B=VD$ as well as previous estimates of $U,D,V$.

1. Solve for $B$ as:

   -  $DB^T = (D^2 + \lambda I)^{-1}D^2U^TY^* = (D^2/(D^2+\lambda))U^TY^* $
   -  $\bar{B}D = \tilde{U}\tilde{D}^2\tilde{V}^T$
      -  $V \leftarrow \tilde{U}$
      -  $D^2 \leftarrow \tilde{D^2}$
   -  $Y^*[\overline\Omega] = AM^T[\overline{\Omega}] = UD^2V^T[\overline \Omega]$
2. Solve for $A$ as:
   -  $DA^T = (D^2 + \lambda I)^{-1}DV^T{Y^*}^T =  (D^2/(D^2+\lambda))(VY^*)^T$
   - $\bar{A}D = \tilde{U}\tilde{D}^2\tilde{V}^T$
      -  $U \leftarrow \tilde{U}$
      -  $D^2 \leftarrow \tilde{D^2}$
   - $Y^*[\overline\Omega] = AM^T[\overline{\Omega}] = UD^2V^T[\overline \Omega]$
3. Compute Frobenious ratio to check for convergence $Frob(U_{old}, D^2_{old}, V_{old}, U, D^2, V)$

4. Iterate over steps 1-3 till convergence.

5. Compute the SVD decomposition  $Y^*V=U D_\sigma^2 V$
   - $D^2 = diag((\sigma_1-\lambda)_+,\cdots,(\sigma_r-\lambda)_+)$
   - $\hat{M} = \hat{Y} = U D^2 V^T$

**************************************************************

Notice that in the calculations above we used $D^2$ and we didn't need to store $D$. Moreover, We don't need to store $A$ and $B$ and we could replace them with $U,D^2,V$. 



The algorithm above provided the same performance as SoftImpute but with a much faster convergence.  Next, we incorporate Covariates into our calculations. We start by ignoring the regularization on the covariate coefficients.

The objective function in  6 becomes as follows:

$$
\begin{aligned}
minimize && \frac{1}{2} \|Y-AB^T - X\beta\|^2_F + \frac{\lambda}{2} (||A||^2_F+||B||^2_F)  && (ALS1)
\end{aligned}
$$
When $A$ is fixed, we estimate both $B$ and $\beta$. Similarly, when $B$ is fixed, we also get new estimates for $A$ and $\beta$.


*****************************************************************

**Algorithm ALS/Covariates**

1. Initialize $A = UD$ where $U_{n_1 \times r}$ is a randomly chosen matrix with orthonormal columns
   and $D = I_r$, the $r \times r$ identity matrix, and $B = VD$ with $V = 0$. Set $Y^*=Y^+=Y$. 
   Get initial estimates for $\hat{\beta} = (X^TX)^{-1}X^TY$.
   Alternatively, any prior solution $\beta$, $A = UD$ and $B = VD$ could be used as a warm start.

2. Repeat till convergence:

   a. Solve for $B$ and $\beta$ as:
   
      -  $DB^T = (D^2 + \lambda I)^{-1}D^2U^TY^+ = (D^2/(D^2+\lambda))U^TY^+$
      -  $\bar{B}D = \tilde{U}\tilde{D}^2\tilde{V}^T$
         -  $V \leftarrow \tilde{U}$
         -  $D^2 \leftarrow \tilde{D^2}$
      - $Y^*[\overline\Omega] = AB^T[\overline{\Omega}] + X \hat\beta[\overline{\Omega}] = UD^2V^T[\overline \Omega] + X \hat \beta[\overline{\Omega}]$
      - $\hat{\beta} = (X^TX)^{-1} X^T Y^*$
      -  $Y^+= Y^* - X \hat \beta$
   b. Solve for $A$ and $\beta$ as:
      -  $DA^T = (D^2 + \lambda I)^{-1}DV^T{Y^+}^T =  (D^2/(D^2+\lambda))(VY^+)^T$
      - $\bar{A}D = \tilde{U}\tilde{D}^2\tilde{V}^T$
         -  $U \leftarrow \tilde{U}$
         -  $D^2 \leftarrow \tilde{D^2}$
      - $Y^*[\overline\Omega] = AB^T[\overline{\Omega}] + X \hat\beta[\overline{\Omega}] = UD^2V^T[\overline \Omega] + X \hat \beta[\overline{\Omega}]$
      - $\hat{\beta} = (X^TX)^{-1} X^T Y^*$
      -  $Y^+ = Y^* - X \hat \beta$
   c. Compute Frobenious ratio to check for convergence $Frob(U_{old}, D^2_{old}, V_{old}, U, D^2, V)$


4. Compute the SVD decomposition  $Y^*V=U D_\sigma^2 V$
   - $D^2 = diag((\sigma_1-\lambda)_+,\cdots,(\sigma_r-\lambda)_+)$
   - $\hat{M} = U D^2 V^T$
   - $\hat{Y} = \hat{B} + X \beta$

**************************************************************

In simpler terms, we could write the algorithm above as:

********************************


**Algorithm ALS/Covariates**

1. Input: Data Matrix $Y$, Covariate Matrix $X$, initial $A,B,\beta$ and $k=0$
2. Output: $(A^*,B^*,\beta^*)$

Repeat Until Convergence:

   1. $k \leftarrow k+1$
   2. $Y^+ \leftarrow P_{\Omega}(Y)+P_{\overline \Omega}(AB^T)- X\beta =  P_{\Omega}(Y -AB^T) + (AB^T-X\beta)$
   3. $A \leftarrow Y^+ B(B^TB+\lambda I)^{-1}$
   4. $Y^* \leftarrow P_{\Omega}(Y)+P_{\overline \Omega}(AB^T+X\beta)$
   5. $\beta = (X^TX)^{-1}X^TY^*$
   6. $Y^+ \leftarrow Y^*- X\beta$
   7. $B^T \leftarrow (A^TA+\lambda I)^{-1}A^TY^+$

**************************************************************



$$
\begin{aligned}
\end{aligned}
$$

